---
title: "London Crystals Moissanite Key" 
subtitle: |
  | STA 6704 
  |
  | Program Name: LondonCrystals_Moissanite_Key.RMD
output:
  pdf_document:
    number_sections: yes
    toc: no
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    df_print: paged
    toc: no
    toc_depth: '3'
---

<!-- \pagenumbering{gobble} -->

# Introduction

Moissanite, a cheaper alternative to diamonds with 
very similar properties and measurements,
is a growing market among Millennials and Gen-Z.
You are a Data Scientist (DS) for London Crystals, 
an international jewelry chain which has over 
\$2 billion in current assets and a significant amount of cash. 
Rather than expanding their diamond operations, 
London Crystals is considering entering
the moissanite business to diversify their sales. 
A large moissanite distributor is open to a
potential acquisition, 
however it has not priced its full inventory yet.
London Crystals has asked for your help to assess 
the value of the potential acquisition. 
You will be working with a subset of the
inventory that is already measured and priced.

The dataset we are investigating here is 
\verb|Moissanite.csv| available on WebCourses. 
The csv file contains a highly modified version 
of the original “diamonds” dataset.

# Work to be Completed

In this assignment, you will be working
to model the data by implementing stepwise 
selection and regularization in order to 
improve upon a multiple linear regression (MLR). 
You begin by creating an MLR model on 
\verb|price| using all available predictors.

# Note

Where-ever instructed, 
you are required to change/edit some parts of the code. 
Then, run the corresponding chunk to answer the questions.

# Preparing the Data

Run the following chunk without changing it to 
perform the necessary data preparation/cleaning.

```{r}
library(caret)

moissanite <- read.csv("Moissanite.csv", stringsAsFactors = TRUE)

moissanite$cut <- factor(moissanite$cut, 
                         levels = c("Fair", "Good", "Very Good", "Premium", "Ideal"))
moissanite$clarity <- factor(moissanite$clarity, 
                             levels = c("I.SI", "VS", "VV", "IF"))

set.seed(161)
ind <- createDataPartition(moissanite$price,
                           list = FALSE, p = 0.7)

train <- moissanite[ind, ]
test <- moissanite[-ind, ]
```


# Exercise 1

You just need to run the following code and 
write your answer in the answer sheet. 

```{r}
mlr.full <- lm(price ~ ., train)

summary(mlr.full)
```
Provide details on the full MLR model’s coefficients. 
List all the variables which are statistically 
insignificant at 5% significance level.

# Exercise 2

Board of Directors at your Company are not interested in model 
interpretability and wants you to focus on optimizing performance. 
Still, you have concerns over the number of variables currently in the model.
You decide to run a stepwise selection procedure 
with selection criterion either AIC or BIC.

Recommend a specific stepwise procedure starting from the full model. 
Justify your recommendation. 
Write down which variables are eliminated and 
also report your final resulting MLR. 

```{r}
library(MASS)

# An empty (i.e., intercept-only) MLR is needed to implement stepAIC.
mlr.mod0 <- lm(price ~ 1, train)

# ----------- Starting Student Code -----------------------#
# Complete the following lines of code to run stepwise selection. 
# stepAIC(mlr.full,
#         k = 2, # Change to log(nrow(train)) for BIC
#         direction = "backward",
#         scope     = list(upper = mlr.full, lower = mlr.mod0))
# ----------- Ending Student Code ------------------------#

# ------------ Starting Key Code -------------------------#
stepAIC(mlr.full,
        k = 2, # Change to log(nrow(train)) for BIC
        direction = "backward",
        scope     = list(upper = mlr.full, lower = mlr.mod0))
# ------------ Ending Key Code ---------------------------#
```

```{r}
# ----------- Starting Student Code -----------------------#
# From the stepAIC output, copy the code that runs the resulting stepwise model.
# mlr.step <- lm(formula = price ~ carat + cut + color + clarity + table + 
#     x + y + z + distribution, data = train)
# ----------- Ending Student Code ------------------------#


# ------------ Starting Key Code -------------------------#
# This is only for answer! 
mlr.step <- lm(formula = price ~ carat + cut + color + clarity + table + 
                 x + y + z + distribution, data = train)
summary(mlr.step)
# ------------ Ending Key Code ---------------------------#
```

# Exercise 3

Your supervisor desires to consider ridge regression.
The provided code specifies candidate lambda values 
ranging from 0.01 to 100,000 for cross-validation; 
do not change this. 
On the other hand, you must provide the 
correct value for the argument alpha. 
Furthermore, your supervisor instructs you to use 
the lambda value given by \$lambda.1se 
(rather than \$lambda.min) for the ridge model.

* List the value of lambda that minimizes the cross-validation error. 

* Describe the effectiveness of ridge regression on the 
data in light of your answer above.


* Explain the impact on the ridge model’s flexibility 
due to your supervisor’s choice of lambda value.

```{r}
library(glmnet)

trainXmat <- model.matrix(price ~ ., data = train)

set.seed(161)
mlr.ridge <- cv.glmnet(x      = trainXmat[, -1],
                       y      = train$price,
                       family = "gaussian",
                       alpha  = 0,  # what's the correct value for ridge???
                       lambda = 10^(seq(-2, 5, by = 0.25)))

plot(mlr.ridge)

mlr.ridge.mod <- glmnet(x      = trainXmat[, -1],
                        y      = train$price,
                        family = "gaussian",
                        lambda = mlr.ridge$lambda.1se,
                        alpha  = 0)  # what's the correct value for ridge???
```

```{r}
mlr.ridge$lambda.min
mlr.ridge$lambda.1se
```

# Exercise 4

Your supervisor also desires to consider lasso regression. 
The provided code specifies candidate lambda values 
ranging from 0.01 to 100,000 for cross-validation; 
do not change this.
On the other hand, you must provide the correct value for the argument alpha. 
Furthermore, your supervisor instructs you to use the lambda value given 
by \$lambda.1se (rather than \$lambda.min) for the lasso model.

Comment on how the coefficients differ between the 
ridge model from Exercise 3 and the lasso model.

```{r}
set.seed(161)
mlr.lasso <- cv.glmnet(x      = trainXmat[, -1],
                       y      = train$price,
                       family = "gaussian",
                       alpha  = 1,  # what's the correct value for lasso???
                       lambda = 10^(seq(-2, 5, by = 0.25)))

plot(mlr.lasso)

mlr.lasso.mod <- glmnet(x      = trainXmat[, -1],
                        y      = train$price,
                        family = "gaussian",
                        lambda = mlr.lasso$lambda.1se,
                        alpha  = 1)  # what's the correct value for lasso???
```

```{r}
coef(mlr.ridge.mod)
coef(mlr.lasso.mod)
```

# Exercise 5

Recommend one of the models to the Board of Directors. 
Justify your recommendation.

```{r}
testXmat <- model.matrix(price ~ ., data = test)

mlr.full.preds <- predict(mlr.full, test)
mlr.step.preds <- predict(mlr.step, test)
mlr.ridge.preds <- predict(mlr.ridge.mod, newx = testXmat[, -1])
mlr.lasso.preds <- predict(mlr.lasso.mod, newx = testXmat[, -1])

print("Training RMSEs")
sqrt(sum((train$price - predict(mlr.full))^2) / nrow(train))
sqrt(sum((train$price - predict(mlr.step))^2) / nrow(train))
sqrt(sum((train$price - predict(mlr.ridge.mod, newx = trainXmat[, -1]))^2) / nrow(train))
sqrt(sum((train$price - predict(mlr.lasso.mod, newx = trainXmat[, -1]))^2) / nrow(train))

print("Test RMSEs")
sqrt(sum((test$price - mlr.full.preds)^2) / nrow(test))
sqrt(sum((test$price - mlr.step.preds)^2) / nrow(test))
sqrt(sum((test$price - mlr.ridge.preds)^2) / nrow(test))
sqrt(sum((test$price - mlr.lasso.preds)^2) / nrow(test))
```

